The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-05-30 source/save.tex

\section{The Convergence of FedAve for General Convex Functions}

\subsection{The Convergence of FedAve for Convex Functions}
In this subsection, we discuss the convergence of FedAve for convex functions,
we consider both smooth (Theorem~\ref{th:sgdcvxsmth}) and non-smooth (Theorem~\ref{th:sgdcvxnonsmoth})
functions.

>>>>

\section{The Convergence of FedAvg for General Convex Functions}

\subsection{The Convergence of FedAvg for Convex Functions}
In this subsection, we discuss the convergence of FedAvg for convex functions,
we consider both smooth (Theorem~\ref{th:sgdcvxsmth}) and non-smooth (Theorem~\ref{th:sgdcvxnonsmoth})
functions.

### 2020-05-30 source/sec_conclude.tex

\section{Conclusions}
This paper provides a comprehensive analysis of the convergence rate of FedAvg
and its accelerated variants in a general federated learning problem with heterogeneous local data and partial participation. We show that both accelerated FedAvg with Nesterov updates and FedAvg
can achieve linear speedup when the number of nodes increases, i.e., $O(1/\sqrt{NT})$
convergence for convex smooth problems and $O(1/NT)$ convergence for strongly 
convex smooth problems. Furthermore, we show that FedAvg can in fact achieve exponential 
convergence for overparameterized strongly convex smooth problems, and we propose the MaSS accelerated Fedavg algorithm, which has provable speedup in convergence rate over
FedAvg on quadratic problems. Last but not least, we empirically
verify the linear speedup of FedAvg and Nesterov accelerated FedAvg for strongly convex, convex smooth, and linear regression problems. The empirical results are well-aligned with our theories.

>>>>

\section{Conclusions}
This paper provides a comprehensive analysis of the convergence rate of FedAvg
and its accelerated variants in a general federated learning problem with heterogeneous local data and partial participation. We show that both accelerated FedAvg with Nesterov updates and FedAvg
can achieve linear speedup when the number of nodes increases, i.e., $O(1/\sqrt{NT})$
convergence for convex smooth problems and $O(1/NT)$ convergence for strongly 
convex smooth problems. [TODO:Add discussion about communication efficiency.]Furthermore, we show that FedAvg can in fact achieve exponential 
convergence for overparameterized strongly convex smooth problems, and we propose the MaSS accelerated Fedavg algorithm, which has provable speedup in convergence rate over
FedAvg on quadratic problems. Last but not least, we empirically
verify the linear speedup of FedAvg and Nesterov accelerated FedAvg for strongly convex, convex smooth, and linear regression problems. The empirical results are well-aligned with our theories.

### 2020-05-30 source/sec_impact.tex

In the meanwhile, the increasing concerns of security and
privacy leaking from both large-companies and the entire society
present significant challenges of proper utilization of local
data. This trade-off between efficiency and privacy or security
lies in the central of many industrial applications, including 
but not limited to recommender systems~\cite{chen2018federated}, virtual assistants~\cite{lamautonomy}, and mobile keyboard prediction~\cite{47586}

In this work, we establish solid theoretical foundation for federated
learning that achieves efficiency and privacy at the same time. 
On one hand, the data is isolated in each local devices, thus prevent
the potential leaking of private information. On the otherhand, the
convergence rates improved substantially if the number of
active local devices increases. In conclusion, our work 
shows a faithful pathway to simultaneously secure people's privacy and improve
users' experience.

>>>>

In the meantime, the increasing concerns of security and
privacy leaking from both large-companies and the entire society
present significant challenges of proper utilization of local
data. This trade-off between efficiency and privacy or security
lies in the center of many industrial applications, including 
but not limited to recommender systems~\cite{chen2018federated}, virtual assistants~\cite{lamautonomy}, and mobile keyboard prediction~\cite{47586}

In this work, we establish solid theoretical foundations for federated
learning that achieves efficiency and privacy at the same time. 
On one hand, the data is isolated in each local devices, thus prevent
the potential leaking of private information. On the otherhand, the
convergence rates improved substantially if the number of
active local devices increases. In conclusion, our work 
shows a faithful pathway to simultaneously secure people's privacy and improve
users' experience.

### 2020-05-30 source/sec_overparameterized.tex

There have been a line of recent works {[}CITE{]} showing that SGD
and accelerated methods achieve exponential convergence in this case,
thanks to the property of ``automatic variance reduction'' {[}CITE{]}.
The natural question is whether such a result also holds in the federated
learning setting. We first show that this is indeed the case and establish
the exponential (linear) convergence of FedAvg with local SGD updates
with constant step size for general strongly convex and smooth overparameterized
problems. In addition, we show that the convergence rate speeds up
linearly in the number of workers $N$ when $N$ is below some problem-dependent
threhold, while decreases with $E$ through $1/E$.

>>>>

There have been a line of recent works {[}CITE{\cite{ma2017power}]} showing that SGD
and accelerated methods achieve exponential convergence in this case,
thanks to the property of ``automatic variance reduction'' {[}CITE{]}.
The natural question is whether such a result also holds in the federated
learning setting. We first show that this is indeed the case and establish
the exponential (linear) convergence of FedAvg with local SGD updates
with constant step size for general strongly convex and smooth overparameterized
problems. In addition, we show that the convergence rate speeds up
linearly in the number of workers $N$ when $N$ is below some problem-dependent
threhold, while decreases with $E$ through $1/E$.

### 2020-05-30 source/sec_relatedwork.tex

% Non-convex linear speedup
In particular, we focus on the convergence analysis of FedAvg, which
has been discussed in various settings. 
In~\cite{yu2019parallel,wang2019adaptive}, the authors provides convergence
rate of FedAvg for non-convex problem, given all the devices are active at
each communication round. Recently, \citep{huo2020faster} provide 
convergence guarantee for Nesterov accelerated FedAvg for non-convex 
setting in partial participation setting. Although the empirical
results is improved over FedAvg, the accelerated version admitting a similar 
convergence rate as FedAvg. In this work, we provide an novel accelerated
FedAvg with improved convergence rate under a popular overparameterized 
setting. In addition, \cite{yu2019linear} provides $O(1/\sqrt{NT})$ 
convergence (i.e., linear speedup) of FedAvg for non-convex problem in the full participation setting. 
For convex smooth problems, \cite{khaled2019first} prove the 
convergence rate of local GD (a full batch version of FedAvg with full participation) on heterogeneous data. 
Concurrently, \cite{li2019convergence} firstly provide exponential convergence of FedAvg for strongly convex problem,
under the more practical setting of partial participation and Non-IID data,
while their results did not achieve linear speedup. 
\cite{stich2018local} prove linear speedup convergence of FedAvg for strongly convex problem, while they assume full participation and i.i.d. data.
Comparing to prior work, our results provide the linear speedup on strongly convex problem in the practical partial participation setting and heterogeneous data.

>>>>

% Non-convex linear speedup
In particular, we focus on the convergence analysis of FedAvg, which
has been discussed in various settings. 
In~\cite{yu2019parallel,wang2019adaptive}, the authors provide the convergence rate of FedAvg for the non-convex problem, given all the devices are active at each communication round. Recently, \citep{huo2020faster} provide 
convergence guarantee for Nesterov accelerated FedAvg for non-convex 
setting in a partial participation setting. Although the empirical
results are improved over FedAvg, the accelerated version admitting a similar convergence rate as FedAvg. In this work, we provide a novel accelerated
FedAvg with improved convergence rates under a popular overparameterized setting. In addition, \cite{yu2019linear} provides $O(1/\sqrt{NT})$ 
convergence (i.e., linear speedup) of FedAvg for the non-convex problem in the full participation setting. 
For convex smooth problems, \cite{khaled2019first} prove the 
convergence rate of local GD (a full batch version of FedAvg with full participation) on heterogeneous data. 
Concurrently, \cite{li2019convergence} firstly provides exponential convergence of FedAvg for the strongly convex problem,
under the more practical setting of partial participation and Non-IID data while their results did not achieve linear speedup. 
\cite{stich2018local} prove linear speedup convergence of FedAvg for the strongly convex problem, while they assume full participation and identically distributed data. Comparing to previous works, our results provide the linear speedup on strongly convex problems in the practical partial participation setting and heterogeneous data. In \cite{liu2019accelerating}, the authors provide a convergence rate of accelerated FedAvg for strongly convex problems, while it is considered in full participation setting. To the best of our knowledge,
we provide the first linear speedup convergence results for accelerated FedAvg in the practical setting.
